Extreme climatic events with asymmetric tail-dependence affect population-extinction risk

### 2019-07-28 MainText.Rmd
writing discussion

<!--how common are the extreme events?-->
For past few decades, there have been increasing evidences that anthropogenic climate change 
is causing extreme events, e.g. temperature or precipitation extremes,
heat-waves, wild-fire etc. [@ummenhofer2017extreme]. Frequent extreme events 
can shift temperature or precipitation-mean, 
change variability of the distribution, as well as change its 
symmetry or skewness. Such extreme climatic events (ECE) can pose severe
impacts at the individual, population, community and ecosystem level [@bragazza2008climatic; 
@gutschick2003extreme; @jiguet2011community; @mckechnie2009climate;
@hoover2014resistance]. Therefore, 
understanding the long-term impacts of changes 
in ECEs has been a longstanding interest in ecological research 
[@katz2005statistics; @smith2011ecological; @diez2012will; @moritz2013future; @bailey2016tackling].

@@

<!--stochastic matrix model : age structured-->
To investigate (Q1) from \nameref{Introduction} that how copula structure in noisy 
environment affect the extinction risk of
populations in a single patch age-structured model,
we considered a basic stochastic age-structured model (Eq. \ref{eq_age_str}) 
with two age classes (egg $E$ and adult $A$). We assumed
the population was semelperous, i.e., adult ($A$) went through the
reproductive episode only once before death. Eggs and adults in the 
next generation depend upon the fecundity of adult ($f_a \geq 0$) 
and survival of eggs ($0 \leq s_e \leq 1$) from present generation, respectively.
The environmental stochasticity in the model was embedded in 
the fluctuating vital rates ($f_a$ and $s_e$). We considered the fecundity ($f_a$)
and survival ($s_e$) rates were random draw from a gamma ($\Gamma(k,\theta)$) and
a beta ($\mathcal{B}(\alpha,\beta)$) distribution, respectively. These distributions 
were generated from iid noises showing specific tail-dependence (either in left- or in right-tail)
between fecundity and survival rates. Tail dependence emerged if extreme bad years ("catastrophs") 
would suppress both vital rates (hence, $f_a$ and $s_e$ would show a left tail dependent copula), whereas 
extreme good years ("bonanzas") would increase both of them (i.e., $f_a$ and $s_e$ would show 
a right tail dependent copula). Eq. \ref{eq_vitalrates} represent the stochasticity 
in the vital rates. We included the density dependence effect in one of the vital rate,
namely, fecundity ($f_a$). If $\Omega = \infty$ in Eq. \ref{eq_vitalrates}, 
the projection matrix would be independent
of populations from any age-class and the model was then called to be "linear", whereas
its "nonlinear" version had finite $\Omega$.

@@

For single patch linear age-structured models, we found left-tail 
dependence (Clayton, Suvival Gumbel copulas) in noisy vital 
rates enhanced the extinction risk compared to the case when
vital rates were right tail dependent (Survival clayton, Gumbel copulas).
It was expected because a sudden "bad" year (e.g. with severe cold) can suppress 
the fecundity of adult and survival/ maturation of eggs simultaneously 
and thereby increasing the extinction risk of the population. For non-linear age-structured 
models, we got the similar trend for extinction risks with different tail-dependence (left and right).
With non-linear model the extinction risks appeared as higher than their corresponding 
linear analog as it was evident from Eq. \ref{eq_vitalrates} with finite $\Omega = 100$. 
Fig. \ref{SM-fig_extrisk_agestr_C_SC_G_SG_linear} represents this scenario with 4 different 
asymmetric copulas but of same Kendall's correlation ($\tau = 0.6$). There was very little
difference between extinction risks within a pair of different tail-dependent copulas
(e.g. C-SC in Fig. \ref{SM-fig_extrisk_agestr_C_SC_G_SG_linear}(A,C) and
G-SG in Fig. \ref{SM-fig_extrisk_agestr_C_SC_G_SG_linear}(B,D)) and that difference was partly 
dependent on the strength of tail dependence. Therefore we repeated the same analysis with 
vital rates coming from extreme-tail dependent (left and right in two different simulations)
copulas and got the similar results but with much prominent difference between their extinction risks
(Fig. \ref{fig_extrisk_agestr}). This result answered (Q1) and showed extinction risk of 
a population could be considered as
direct manifestation of tail-dependence measure between vital rates of different age-class
within that population. <!--discussion later : add Tuljapurkar's approximation-->

@@

<!--Tuljapurkar approximation : linear version of age-structured model in stochastic environments-->

<!--future study: worthy for PVA, IUCN red list
1. Identify which life stage or demographic process as management target?
2. multisite age str. model with dispersal, directional change in spatio-temporal variation for vital rates.
3. Old study (Doak1994, Ferguson1995) showed including correlation btw vital rates for PVA makes a difference, 
   what if we include TD through copula?
4. What if TD with negative correlation?
e.g. some events that will kill adults will allow more successful recruitment of seeds or other propagules.
-->

>>>>

<!--how common are the extreme events?-->
For past few decades, there have been increasing evidences that anthropogenic climate change 
is causing extreme events, e.g. temperature or precipitation extremes,
heat-waves, wild-fire etc. [@ummenhofer2017extreme]. Frequent extreme events 
can shift temperature or precipitation-mean, 
change variability of the distribution, as well as change its 
symmetry or skewness [@hansen2012perception]. Such extreme climatic events (ECE) can pose severe
impacts at the individual, population, community and ecosystem level [@bragazza2008climatic; 
@gutschick2003extreme; @jiguet2011community; @mckechnie2009climate;
@hoover2014resistance]. Therefore, 
understanding the long-term impacts of changes 
in ECEs has been a longstanding interest in ecological research 
[@katz2005statistics; @smith2011ecological; @diez2012will; @moritz2013future; @bailey2016tackling].

@@

<!--stochastic matrix model : age structured-->
To investigate (Q1) from \nameref{Introduction} that how copula structure in noisy 
environment affect the extinction risk of
populations in a single patch age-structured model,
we considered a basic stochastic age-structured model (Eq. \ref{eq_age_str}) 
with two age classes (egg $E$ and female adult $A$). We assumed
the population was semelperous, i.e., adult ($A$) went through the
reproductive episode only once before death. Eggs and adults in the 
next generation depend upon the fecundity of adult ($f_a \geq 0$) 
and survival of eggs ($0 \leq s_e \leq 1$) from present generation, respectively.
The environmental stochasticity in the model was embedded in 
the fluctuating vital rates ($f_a$ and $s_e$). We considered the fecundity ($f_a$)
and survival ($s_e$) rates were random draw from a gamma ($\Gamma(k,\theta)$) and
a beta ($\mathcal{B}(\alpha,\beta)$) distribution, respectively. These distributions 
were generated from iid noises showing specific tail-dependence (either in left- or in right-tail)
between fecundity and survival rates. Tail dependence emerged if extreme bad years ("catastrophs") 
would suppress both vital rates (hence, $f_a$ and $s_e$ would show a left tail dependent copula), whereas 
extreme good years ("bonanzas") would increase both of them (i.e., $f_a$ and $s_e$ would show 
a right tail dependent copula). Eq. \ref{eq_vitalrates} represent the stochasticity 
in the vital rates. We included the density dependence effect in one of the vital rate,
namely, fecundity ($f_a$). If $\Omega = \infty$ in Eq. \ref{eq_vitalrates}, 
the projection matrix would be independent
of populations from any age-class and the model was then called to be "linear", whereas
its "nonlinear" version had finite $\Omega$.

@@

For single patch linear age-structured models, we found left-tail 
dependence (Clayton, Suvival Gumbel copulas) in noisy vital 
rates enhanced the extinction risk compared to the case when
vital rates were right tail dependent (Survival clayton, Gumbel copulas).
It was expected because a sudden "bad" year (e.g. with severe cold) can suppress 
the fecundity of adult and survival/ maturation of eggs simultaneously 
and thereby increasing the extinction risk of the population. For non-linear age-structured 
models, we got the similar trend for extinction risks with different tail-dependence (left and right).
With non-linear model the extinction risks appeared as higher than their corresponding 
linear analog as it was evident from Eq. \ref{eq_vitalrates} with finite $\Omega = 100$. 
Fig. \ref{SM-fig_extrisk_agestr_C_SC_G_SG} represents this scenario with 4 different 
asymmetric copulas but of same Kendall's correlation ($\tau = 0.6$). There was very little
difference between extinction risks within a pair of different tail-dependent copulas
(e.g. C-SC in Fig. \ref{SM-fig_extrisk_agestr_C_SC_G_SG}(A,C) and
G-SG in Fig. \ref{SM-fig_extrisk_agestr_C_SC_G_SG}(B,D)) and that difference was partly 
dependent on the strength of tail dependence. Therefore we repeated the same analysis with 
vital rates coming from extreme-tail dependent (left and right in two different simulations)
copulas and got the similar results but with much prominent difference between their extinction risks
(Fig. \ref{fig_extrisk_agestr}). This result answered (Q1) and showed extinction risk of 
a population could be considered as
direct manifestation of tail-dependence measure between vital rates of different age-class
within that population. <!--discussion later : add Tuljapurkar's approximation-->

@@

<!--Tuljapurkar approximation : linear version of age-structured model in stochastic environments-->
Though we found left tail dependence in vital rates (induced by ECEs) caused higher extinction risk than 
the right tail dependent ones, the difference was not remarkable (Fig. \ref{SM-fig_extrisk_agestr_C_SC_G_SG}) 
unless the tail dependence was at its extreme (Fig. \ref{fig_extrisk_agestr}). This finding could be explained 
by the approximation @tuljapurkar1990lecture made for the stochastic log(growth rate) in an iid environment.
The approximation considered that the growth rate depends on the most dominant eigen value of the mean projection matrix 
and the temporal variance of the log(population growth rate) caused by environmental stochasticity, which was based
on the linear covariance measure between any two matrix elements, not on their tail-dependence properties. Therefore,
according to Tuljapurkar's approximation, population growth rate in an iid environment 
remained same for our model (Eq. \ref{eq_age_str}) with left-tail or right-tail dependent noises.

<!--future study: worthy for PVA, IUCN red list
1. Identify which life stage or demographic process as management target?
2. Old study (Doak1994, Ferguson1995) showed including correlation btw vital rates for PVA makes a difference, 
   what if we include TD through copula?
3. multisite age str. model with dispersal, directional change in spatio-temporal variation for vital rates.
-->
Although we demonstrated that tail dependence between vital rates of an age structured model can
influence extinction risk, substantial work remains to determine the importance of this effect in the 
following scenarios. First, we demonstrated the effect of tail-dependence on extinction risk 
with a very simplistic age structured (two age stages) model where we considered the total semelperous 
population would go extinct if there would be only one female adult and one egg. But in reality population 
viability analysis (PVA) deals with a little bit higher threshold with several age, stage or size class and a variety 
of populations including semelperous and iteroperous [@morris2002quantitative]. Future studies should include 
tail-dependence effect on extinction risk 
for a multiple age-, size- and stage- structured model and estimate extinction risk 
for each age, size and stage class separately 
so that we can identify (via multisite-PVA) which life-stage or demographic process 
to target for population management. Earlier study by @doak1994modeling showed PVA models including correlation 
between different demographic rates indicated marked difference in predicting confidence interval for 
future population size than the model which did not include correlation. But when 
such correlation cannot be estimated directly, more generally where the
forms of statistical dependence among the model's parameters are unknown, making incorrect
assumptions about correlations and dependencies may lead to severe over-estimation 
or under-estimation of extinction risks [@ferson1995correlations]. We believe modeling 
dependencies with copula could help in this case. Second, we did not consider multiple sites of 
age-structured metapopulation which could be linked via dispersal. So, possible future investigation 
would be to explore the viability for the metapopulations if vital rates across sites 
showed specific tail dependence with spatially correlated extreme climatic events.

<!--What if TD with negative correlation?
e.g. some events that will kill adults will allow more successful recruitment of seeds or other propagules.
-->
